Combining Static and Dynamic Analyses to Secure Programs against Integer Overflows

### 2012-08-31 doc/Manuscript/CGO/paper.tex
Added new figs.
Added more text to Raphael's section.

%What is our instrumentation library
We have implemented our instrumentation library as a LLVM transformation pass;
thus, we work at the level of the compiler's intermediate
representation~\footnote{\texttt{http://llvm.org/docs/LangRef.html}}.
This is in contrast to previous work, which either transforms the
source code~\cite{Dietz12}, or the machine dependent code~\cite{Brumley07}.
We work at the intermediate representation level to be able to couple our
library with static analyses, such as the algorithm that we described in
Section~\ref{sec:range}.
Our instrumentation works by identifying the instructions that may 
lead to an overflow, and inserting assertions after those instructions. 
The LLVM IR has five instructions that may lead to an overflow:
\textsc{Add}, \textsc{Sub}, \textsc{Mul},  \textsc{Trunc} and \textsc{Shl}
(left shift).
Figure~\ref{fig:instrumentation} shows the dynamic tests that we perform to
detect overflows.
This figure also shows how the static analysis helps us to avoid inserting
some of those tests.
We only generate the dynamic tests on the right if the conditions in the middle
are false.

\begin{figure*}[t!]
\begin{center}
\begin{tabular}{lll}
Instruction & Static Check & Dynamic Check \\ \\
$x = o_1 \ +_s \ o_2$ & $\lb{I[o_1]} + \lb{I[o_2]} > \mbox{MIN\_INT} \wedge \ub{I[o_1]} + \ub{I[o_2]} < \mbox{MAX\_INT}$ & $(o_1 > 0 \wedge o_2 > 0 \wedge x < 0) \ \ \vee$ \\
                  & & $(o_1 < 0 \wedge o_2 < 0 \wedge x > 0)$ \\ \\
$x = o_1 \ +_u \ o_2$ & $\ub{I[o_1]} + \ub{I[o_2]} < \mbox{MAX\_INT}$ & $x < o_1 \vee x < o_2$ \\ \\
$x = o_1 \ -_s \ o_2$ & $\lb{I[o_1]} - \lb{I[o_2]} > \mbox{MIN\_INT} \wedge \ub{I[o_1]} - \ub{I[o_2]} < \mbox{MAX\_INT}$ & $(o_1 < 0 \vee o_2 > 0 \vee x > 0) \ \ \vee$ \\
                      & & $(o_1 > 0 \vee o_2 < 0 \vee x < 0)$ \\ \\
$x = o_1 \ -_u \ o_2$ & $\lb{I[o_1]} > \ub{I[o_2]}$ & $o_1 < o_2$ \\ \\
$x = o_1 \ \times_{u/s} \ o_2$ & $\mbox{MIN\_INT} < \{ \lb{I[o_1]} \times \lb{I[o_2]}, \lb{I[o_1]} \times \ub{I[o_2]} \}< \mbox{MAX\_INT}$ & $x \neq 0 \Rightarrow x \div o_1 \neq o_2$ \\
                       & $\mbox{MIN\_INT} < \{ \ub{I[o_1]}\times \lb{I[o_2]}, \ub{I[o_1]}\times \ub{I[o_2]} \}< \mbox{MAX\_INT}$ & \\ \\
$x = o_1 \ \qside \ n$ & $2^n \times \ub{I[o_1]} < \mbox{MAX\_INT}$ & $(o_1 > 0 \wedge x < o_1) \vee (o_1 < 0 \wedge n \neq 0)$ \\ \\
$x = \ \downarrow_n \ o_1$  & $\lb{I[o_1]} > -2^n \wedge \ub{I[o_1]} < 2^n_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\ \\
$x = \ (\mbox{type}_n) o_1$ & $\lb{I[o_1]} > -2^n \wedge \ub{I[o_1]} < 2^n_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\
\end{tabular}
\end{center}
\caption{\label{fig:instrumentation}Overflow checks. We use $\downarrow_n$ for
the operation that truncates to $n$ bits, and we use $(\mbox{type}_n)$ for a
bit-cast to $n$ bits. Both operations have the same semantics in our setting.
The subscript $s$ indicates a signed instruction; the subscript $u$ indicate
an unsigned operation. If the static check is true, we do not insert the dynamic
check. The values \{MAX/MIN\}\_INT depend on the type of the instruction.}
\end{figure*}

%SUB instrumentation
The instrumentation that we insert is mostly straightforward.
We discuss in the rest of this section a few interesting cases.
If we are dealing with an unsigned \textsc{Sub} instruction, e.g,
$x = o_1 \ -_u \ o_2$, then a single check is enough to detect the bad
behavior: $o_1 < o_2$.
This is sound because if $o_2$ is greater than $o_1$, then we assume that it is
a bug to try to represent a negative number in unsigned arithmetics.
Regarding multiplication, e.g., $x = o_1 \times o_2$, if $o_1 = 0$, then this
operation can never cause an overflow.
This test is necessary to us, because we check integer overflows in
multiplication via the inverse operation of integer division.
Thus, the test prevents a division by zero from happening.
The \textsc{Trunc} instruction, e.g., $x = \ \downarrow_n \ o_1$ assigns to $x$
the $n$ least significant bits of $o_1$.
The dynamic checking in this case consists in expanding $x$ to
the datatype of $o_1$ and comparing the expanded value with $o_1$.
The LLVM IR provides instructions to perform these type expansions.
Note that our instrumentation catches any truncation that might result in
data loss, even if this loss is benign.
To make the dynamic checks more liberal, we give users the possibility of
disabling tests over truncations.

\paragraph{Practical Considerations: }
%Overview about the instrumentation technique
Our instrumentation library works by adding new instructions to
the target program.
Although the dynamic check depends on the instruction that is instrumented,
the general modus operandi is the same.
Our dynamic tests check for overflows after they happen.
The code that we insert to detect the overflow diverts the program flow in case
such an event takes place.

% TODO: add a snapshot of a function with, and without the integer overflow
% instrumentation. Let this function be very simple, e.g.,
% int foo(int x, int y) {return x + y; }

Which actions are performed once the overflow is detected depends on the
user.
Our library gives the user three options to handle them. 
First option: no-op.
This option allows us to verify the slowdown produced by the new instructions.
Second option: logging.
This is the standard option, and it preserves the behavior of the instrumented
program.
Whenever an overflow is detected, we print \texttt{Overflow detected in
FileName.cpp, line X.} in the standard error stream.
Third option: abort.
This option terminates the program once an overflow is detected.
Thus, it disallows undefined behavior due to integer overflows.

>>>>

%What is our instrumentation library
We have implemented our instrumentation library as a LLVM transformation pass;
thus, we work at the level of the compiler's intermediate
representation~\footnote{\texttt{http://llvm.org/docs/LangRef.html}}.
This is in contrast to previous work, which either transforms the
source code~\cite{Dietz12}, or the machine dependent code~\cite{Brumley07}.
We work at the intermediate representation level to be able to couple our
library with static analyses, such as the algorithm that we described in
Section~\ref{sec:range}.
Our instrumentation works by identifying the instructions that may 
lead to an overflow, and inserting assertions after those instructions. 
The LLVM IR has five instructions that may lead to an overflow:
\textsc{Add}, \textsc{Sub}, \textsc{Mul},  \textsc{Trunc} and \textsc{Shl}
(left shift).
Figure~\ref{fig:instrumentation} shows the dynamic tests that we perform to
detect overflows.

\begin{figure}[t!]
\begin{center}
\begin{tabular}{ll}
Instruction & Dynamic Check \\ \\
$x = o_1 \ +_s \ o_2$ & $(o_1 > 0 \wedge o_2 > 0 \wedge x < 0) \ \ \vee$ \\
                      & $(o_1 < 0 \wedge o_2 < 0 \wedge x > 0)$ \\ \\
$x = o_1 \ +_u \ o_2$ & $x < o_1 \vee x < o_2$ \\ \\
$x = o_1 \ -_s \ o_2$ & $(o_1 < 0 \vee o_2 > 0 \vee x > 0) \ \ \vee$ \\
                      & $(o_1 > 0 \vee o_2 < 0 \vee x < 0)$ \\ \\
$x = o_1 \ -_u \ o_2$ & $o_1 < o_2$ \\ \\
$x = o_1 \ \times_{u/s} \ o_2$ & $x \neq 0 \Rightarrow x \div o_1 \neq o_2$ \\ \\
$x = o_1 \ \qside \ n$ & $(o_1 > 0 \wedge x < o_1) \vee (o_1 < 0 \wedge n \neq 0)$ \\ \\
$x = \ \downarrow_n \ o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\ \\
$x = \ (\mbox{type}_n) o_1$ & $\mbox{cast}(x, \mbox{type}(o_1)) \neq o_1$ \\
\end{tabular}
\end{center}
\caption{\label{fig:instrumentation}Overflow checks. We use $\downarrow_n$ for
the operation that truncates to $n$ bits, and we use $(\mbox{type}_n)$ for a
bit-cast to $n$ bits. Both operations have the same semantics in our setting.
The subscript $s$ indicates a signed instruction; the subscript $u$ indicate
an unsigned operation.}
\end{figure}

%SUB instrumentation
The instrumentation that we insert is mostly straightforward.
We discuss in the rest of this section a few interesting cases.
If we are dealing with an unsigned \textsc{Sub} instruction, e.g,
$x = o_1 \ -_u \ o_2$, then a single check is enough to detect the bad
behavior: $o_1 < o_2$.
This is sound because if $o_2$ is greater than $o_1$, then we assume that it is
a bug to try to represent a negative number in unsigned arithmetics.
Regarding multiplication, e.g., $x = o_1 \times o_2$, if $o_1 = 0$, then this
operation can never cause an overflow.
This test is necessary, because we check integer overflows in
multiplication via the inverse operation, e.g., integer division.
Thus, the test prevents a division by zero from happening.
The \textsc{Trunc} instruction, e.g., $x = \ \downarrow_n \ o_1$ assigns to $x$
the $n$ least significant bits of $o_1$.
The dynamic check, in this case, consists in expanding $x$ to
the datatype of $o_1$ and comparing the expanded value with $o_1$.
The LLVM IR provides instructions to perform these type expansions.
Note that our instrumentation catches any truncation that might result in
data loss, even if this loss is benign.
To make the dynamic checks more liberal, we give users the possibility of
disabling tests over truncations.

\paragraph{Practical Considerations: }
%Overview about the instrumentation technique
Our instrumentation library works by adding new instructions to
the target program.
Although the dynamic check depends on the instruction that is instrumented,
the general modus operandi is the same.
The dynamic tests check for overflows after they happen.
The code that we insert to detect the overflow diverts the program flow in case
such an event takes place.
Figure~\ref{fig:instrumented_cfg} shows an actual control flow graph, before
and after the instrumentation.
From this figure, it is clear that the instrumented program will be larger
than the original code.
However, as we show empirically, this growth is not too large when compared
to the total size of our benchmarks, because most of the instructions in these
programs do not demand instrumentation.
Furthermore, none of the instructions used to instrument integer arithmetics
access memory.
Therefore, the overall slowdown that the instrumentation causes is usually
small.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\columnwidth]{images/instrumented_cfg}
\end{center}
\caption{\label{fig:instrumented_cfg}
(a) A simple C function.
(b) The same function converted to the LLVM intermediate representation.
(c) The instrumented code. The boldface lines were part of the original
program.}
\end{figure}

Which actions are performed once the overflow is detected depends on the
user, who has the option to overwrite the \texttt{handle\_overflow}
function in Figure~\ref{fig:instrumented_cfg}.
Our library gives the user three options to handle them. 
First option: no-op.
This option allows us to verify the slowdown produced by the new instructions.
Second option: logging.
This is the standard option, and it preserves the behavior of the instrumented
program.
Whenever an overflow is detected, we print \texttt{Overflow detected in
FileName.cpp, line X.} in the standard error stream.
Third option: abort.
This option terminates the program once an overflow is detected.
Thus, it disallows undefined behavior due to integer overflows.

\paragraph{Using the static analysis to avoid some overflow checks.}
Our library can, optionally, use the range analysis to avoid having to
insert some overflow checks into the instrumented program.
We give the user the possibility of calling the range analysis with either the
e-SSA or the u-SSA live range splitting strategies.
Our static analysis classifies variables into four categories, depending on
their bounds:
\begin{itemize}
\item \textbf{Safe}: a variable is safe if its bounds are fully contained
inside its declared type.
For instance, if $x$ is declared as an unsigned 8-bits integer, then $x$ is
safe if its bounds are within the interval $[0, 255]$.
\item \textbf{Suspicious}: we say that a variable is suspicious if its bounds
go beyond the interval of its declared type, but the intersection between
these two ranges is non-empty.
For instance, the same variable $x$ would be suspicious if
$I[x] = [0, 257]$, as $\ub{I[x]} > \ub{\mathtt{int8}}$.
\item \textbf{Uncertain}: we classify a variable as uncertain if at least one
of its limits is unbounded.
Our variable $x$ would be uncertain if $I[x] = [0, \infty]$.
We distinguish suspicious from uncertain variables because we speculate that
actual overflows are more common among elements in the former category.
\item \textbf{buggy}: a variable is buggy if the intersection between its
inferred range and the range of its declared type is empty.
This is a definitive case of an overflow.
Continuing with our example, $x$ would be buggy if, for instance,
$I[x] = [257, \infty]$, given that $[257, \infty] \sqcap [0, 255] = \emptyset$.
\end{itemize}
Independent on the arithmetic instruction that is being analyzed, the
instrumentation library performs the same test:
if the result $x$ of an arithmetic instruction such as $x = o_1 +_s \ o_2$ is
safe, then the overflow check is not necessary, otherwise it must be
created.
